---
title: "Assemble grouping traits and form Functional Entities"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(oharac)
oharac::setup()
source(here('fb_slb_fxns.R'))


```

# Summary

Gather processed traits data and assemble into a dataframe.  Use trait bins to identify functional entities (FE) per Villeger and others.

# Data

See individual trait scripts.

# Methods

## Gather traits

Here we will read in each trait set and assign the values into specific bins as appropriate.  Inspired by Sebastien Villeger's code, I will create text codes for each trait/value combo, but with a few more letters so they are a little more comprehensible.

### mobility traits

```{r mobility}
mob_codes <- c('sessile'           = 'ses',
               'sedentary'         = 'sed',
               'mobile resident'   = 'res',
               'migratory/nomadic' = 'mig')
mob_traits <- read_csv(here('_data/traits_grouping/trait_mobility.csv')) %>%
  mutate(mob = mob_codes[value],
         mob = factor(mob, levels = mob_codes, ordered = TRUE)) %>%
  group_by(species) %>%
  filter(mob == max(mob)) %>%
  ungroup() %>%
  select(species, mob) %>%
  distinct()

# x <- show_dupes(mob_traits, 'species')

table(mob_traits$mob) %>% knitr::kable()
```

### water column position traits

```{r water col position}
trait_val_codes <- c(reef    = 'rf',
                     pelagic = 'pel',
                     benthic = 'ben',
                     benthopelagic = 'bp')
water_col_traits <- read_csv(here('_data/traits_grouping/trait_water_col_position.csv')) %>%
  mutate(wcol = trait_val_codes[value],
         wcol = factor(wcol, levels = trait_val_codes, ordered = FALSE)) %>%
  select(species, wcol) %>%
  distinct()

# x <- show_dupes(water_col_traits, 'species')


table(water_col_traits$wcol) %>% knitr::kable()
```

### Reproduction traits

Here we have options of:

* generation time
* fecundity
* parental investment
* planktonic larval duration
* post-birth/hatching parental dependence

Generation time and fecundity seem most important, but others can be included as separate traits as well.  For spp with multiple fecundity and generation time values, take the largest category.

#### Fecundity

```{r fecundity}
repr_traits <- read_csv(here('_data/traits_grouping/trait_reproduction.csv')) %>%
  filter(!is.na(value))

fecund_codes <- c('<1'         = 'lo',
                  '1-2'        = 'lo',
                  '2-5'        = 'lo',
                  '5-10'       = 'lo',
                  '10-20'      = 'med',
                  '20-50'      = 'med',
                  '50-100'     = 'med',
                  '100-1000'   = 'hi',
                  '1000-10000' = 'vhi',
                  '>10000'     = 'vhi')

fecund_traits <- repr_traits %>%
  filter(trait == 'fec') %>%
  mutate(fec = fecund_codes[value],
         fec = factor(fec, levels = unique(fecund_codes), ordered = TRUE)) %>%
  group_by(species) %>%
  filter(fec == max(fec)) %>%
  ungroup() %>%
  select(species, fec) %>%
  distinct()

table(fecund_traits$fec)
```

#### Generation time

```{r gen time}

gen_codes <- c("<1yr"     = 'vshort',
               "1-5yrs"   = 'short',
               "5-10yrs"  = 'med',
               "10-20yrs" = 'long',
               ">20yrs"   = 'long')

gen_time_traits <- repr_traits %>%
  filter(trait == 'age_mat') %>%
  mutate(gen = gen_codes[value],
         gen = factor(gen, levels = unique(gen_codes), ordered = TRUE)) %>%
  group_by(species) %>%
  filter(gen == max(gen)) %>%
  ungroup() %>%
  select(species, gen) %>%
  distinct()

table(gen_time_traits$gen)
```

#### check correlation between gen time and fecundity 

```{r}
gen_fec_df <- full_join(fecund_traits, gen_time_traits, by = 'species') %>%
  drop_na()

gen_fec_mtx <- gen_fec_df %>%
  group_by(gen, fec) %>%
  summarize(n = n()) %>%
  mutate(fec = paste0('f:', fec),
         gen = paste0('g:', gen)) %>%
  spread(fec, n) %>%
  column_to_rownames('gen') %>%
  as.matrix()

chi2 <- chisq.test(gen_fec_mtx, correct=F)
c(chi2$statistic, chi2$p.value)

ggplot(gen_fec_df, aes(x = gen, y = fec)) +
  geom_jitter(alpha = .2)
```

While there is obviously correlation, there seems to be enough variation to keep both variables in for now.

### Trophic level

Similar to Mireia's classifications: classify into primary producers (1), primary consumers (herbivores/detritivores) 2), secondary (2-3), tertiary (3-4), apex (>4).  Examine potential break points to see distributions.

Some considerations:

* _Orcinus orca_ is 4.6
* _Carcharodon carcharias_ is 4.48
* _Oncorhynchus_ spp are from 3.23 (Sockeye) to 4.18 (Coho, the only one above 4.0)

Here we set break points at the top value for each category ($\leq$ 2 = primary consumer, $\leq$ 3 = secondary consumer, $\leq$ 4 = tertiary consumer, >4 = apex consumer)

```{r trophic level}
troph_traits <- read_csv(here('_data/traits_grouping/trait_trophic_level.csv')) %>%
  filter(!is.na(value))

summary(troph_traits)

troph_qtile <- ecdf(troph_traits$value)
troph_qtile(c(2, 3, 3.5, 4, 4.4, 5))

troph_traits <- troph_traits %>%
  mutate(trp = case_when(value == 2 ~ 'pri', ### anything above 2 suggests some carnivory
                         value <= 3 ~ 'sec',
                         value <= 4 ~ 'ter',
                         value > 4  ~ 'apex',
                         TRUE ~ 'oops'),
         trp = factor(trp, levels = c('pri', 'sec', 'ter', 'apex'), ordered = TRUE)) %>%
  select(species, trp) %>%
  distinct()

table(troph_traits$trp)
```

### Max length

Examine distribution of lengths to see if there are good break points, or choose break points to achieve a reasonable distribution across reasonable size classes (~ factor of 10 for each class).

```{r max length}
length_traits <- read_csv(here('_data/traits_grouping/trait_length_cm.csv')) %>%
  filter(!is.na(value))

summary(length_traits$value)
quantile(log10(length_traits$value), c(.2, .4, .6, .8))

length_traits <- length_traits %>%
  mutate(len = case_when(value <    1.5 ~ 'vsm', ### smaller than 1.5 cm
                         value <    5.0 ~ 'sm',  ### 1.5 - 5 cm
                         value <   15.0 ~ 'med', ### 5 - 15 cm
                         value <  150.0 ~ 'lg',  ### 15 - 150 cm
                         value >= 150.0 ~ 'vlg', ### larger than 1.5 meters
                         TRUE ~ 'oops'),
         len = factor(len, levels = c('vsm', 'sm', 'med', 'lg', 'vlg'), ordered = TRUE)) %>%
  select(species, len) %>%
  distinct()

table(length_traits$len)
```

The number of uniquely identified species with full complement of traits will be limited by data availability in the Species Vulnerability project as well as data availability in Fishbase/Sealifebase.  See individual scripts for various gapfilling and assignation of traits for each distinct trait set.

## Assemble and assign to functional entities

Here we will full-join all trait dataframes by species, eliminate any with NA values for any traits, and then collapse the trait values together into a specific functional entity character code (which will also be assigned a numeric FE code).

```{r}
n_fe_possible <- n_distinct(mob_traits$mob) *
                 n_distinct(fecund_traits$fec) *
                 n_distinct(gen_time_traits$gen) *
                 n_distinct(troph_traits$trp) *
                 n_distinct(length_traits$len) *
                 n_distinct(water_col_traits$wcol)
```

Based on the given trait categories, we could see up to `r n_fe_possible` functional entities, though most will not be populated (e.g., not so likely to find a very large species with a high fecundity and short generation time).

### Number of spp with NAs for one or more traits

``` {r na_count}
all_traits_df <- mob_traits %>%
  full_join(fecund_traits, by = 'species') %>%
  full_join(gen_time_traits, by = 'species') %>%
  full_join(troph_traits, by = 'species') %>%
  full_join(length_traits, by = 'species') %>%
  full_join(water_col_traits, by = 'species') %>%
  left_join(get_worms())

fe_nas <- all_traits_df %>%
  gather(trait, val, -species) %>%
  group_by(species) %>%
  summarize(n_na = sum(is.na(val)))
table(fe_nas$n_na)
```

Which traits are responsible for dropping most species?  examine species dropped for one or two NAs.

``` {r check which traits are causing drops}
traits <- c('mob', 'fec', 'gen', 'trp', 'len', 'wcol')
fe_trait_single_na <- all_traits_df %>% gather(trait, val, all_of(traits)) %>%
  group_by(species) %>%
  filter(is.na(val)) %>%
  filter(n() == 1) %>%
  ungroup()
table(fe_trait_single_na$trait)

fe_trait_double_na <- all_traits_df %>% gather(trait, val, all_of(traits)) %>%
  group_by(species) %>%
  filter(is.na(val)) %>%
  filter(n() == 2) %>%
  arrange(trait) %>%
  summarize(t1 = first(trait), t2 = last(trait)) %>%
  ungroup()
table(fe_trait_double_na %>% select(t1, t2))

fe_trait_triple_na <- all_traits_df %>% gather(trait, val, all_of(traits)) %>%
  group_by(species) %>%
  filter(is.na(val)) %>%
  filter(n() == 3) %>%
  arrange(trait) %>%
  summarize(t1 = first(trait), t2 = nth(trait, 2), t3 = last(trait)) %>%
  ungroup()
table(fe_trait_triple_na %>% select(t1, t2, t3))

```

### assign FE codes

```{r}
fe_df <- all_traits_df %>%
  drop_na() %>%
  mutate(fe_code = sprintf('MOB%s_FEC%s_GEN%s_TRP%s_LEN%s_WCOL%s', mob, fec, gen, trp, len, wcol)) %>%
  group_by(fe_code) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  arrange(desc(n)) %>%
  mutate(fe_code = fct_inorder(fe_code))
```

``` {r examine distributions of species across functional entities}

fe_df_summary <- fe_df %>%
  group_by(fe_code) %>%
  summarize(n_spp = n_distinct(species)) 

hist(log10(fe_df_summary$n_spp))
DT::datatable(fe_df_summary)
```

## Write out functional entities

```{r}
fe_out_df <- fe_df %>%
  select(species, fe_code, n_spp = n) %>%
  mutate(fe_id = as.integer(fe_code))

write_csv(fe_out_df, here('_output/functional_entities/species_fe.csv'))

fe_traits_df <- fe_df %>%
  select(fe_code, n_spp = n, mob, fec, gen, trp, len, wcol) %>%
  distinct()

write_csv(fe_traits_df, here('_output/functional_entities/fe_traits.csv'))

```

