---
title: "Calculate functional diversity metrics"
author: "Casey O'Hara"
date: "7/25/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 7)

library(raster)
library(oharac)
oharac::setup()
source(here('fb_slb_fxns.R'))
```

# Summary

Using functional entities defined in a prior script based on a suite of categorical traits, in combination with spatial distribution of species from AquaMaps, calculate functional diversity metrics based on Mouillot et al (2014) and code from Sebastien Villeger: functional richness, functional vulnerability, functional redundancy, and functional over-redundancy.

# Methods

## Load data

```{r load data}

fe_metrics_file <- here('_output/functional_entities', 'fe_metrics.csv')
# unlink(fe_metrics_file)
if(!file.exists(fe_metrics_file)) {
  am_to_worms_fixes <- read_csv(here('_raw/aquamaps_to_worms_fixes.csv'))
  spp_fe <- read_csv(here('_output/functional_entities/species_fe.csv')) %>%
    left_join(am_to_worms_fixes, by = c('species' = 'species_worms')) %>%
    mutate(species_am = ifelse(is.na(species_am), species, species_am)) %>%
    select(-am_sid, -source)
  am_spp_cells <- get_am_spp_cells()
  am_spp_info  <- get_am_spp_info() %>%
    select(-species, species = sciname)
}
```

## Identify AquaMaps species included in Functional Entity calculations

Then filter the species-cell lookup to just those species, for better performance.

```{r join FEs to aquamaps}
if(!file.exists(fe_metrics_file)) {
  am_spp_fe   <- inner_join(am_spp_info, spp_fe, by = c('species' = 'species_am')) %>%
    select(am_sid, species, occur_cells, iucn_sid, fe_code, fe_id, n_spp)
  
    
  am_cells_fe <- oharac::dt_join(am_spp_fe, am_spp_cells, by = 'am_sid', type = 'left') %>%
    select(loiczid, am_sid, fe_code, fe_id, prob)
}
```

## Calculate functional diversity metrics 

This set of calculations replicates the results of Villeger's function but without resorting to a presence/absence matrix, instead just relying on the original AquaMaps species-cell table.

* spp_fe_df joined to am_spp_cell; 
* group by cell and functional group;
* calculate number of spp per functional group;
* group by cell;
* calculate functional entity metrics - 
    * Species richness
    * Functional entity richness
    * Functional vulnerability (% of FEs represented by a single spp)
    * Weighted functional vulnerability (new: vuln of an FE is $\frac{1}{n_{spp}}$, take average over all FE)
    * Functional redundancy (mean spp count per FE)
    * Functional over-redundancy (sum of spp per FE above mean, divided by total number of spp)
    

```{r calc fe metrics}

if(!file.exists(fe_metrics_file)) {
  # system.time({
  
    am_cells_fe_thresh <- am_cells_fe %>%
      filter(prob >= .4) %>%
      select(-prob) %>%
      distinct()
    
    # system.time({
      nspp_df <- am_cells_fe_thresh %>%
        group_by(loiczid) %>%
        summarize(n_spp = n_distinct(am_sid),
                  n_fe  = n_distinct(fe_code),
                  .groups = 'drop')
    # }) ### 17 seconds
    
    ### this one is a bit slow:
    # system.time({
      nspp_fe_df <- am_cells_fe_thresh %>%
        group_by(loiczid, fe_code) %>%
        summarize(nspp_fe = n(), .groups = 'drop')
          ### using n() instead of n_distinct(); if no dupes
          ### in am_sid/loiczid/fe_code then should be fine.
    # }) ### 175 seconds
    
    # system.time({
      f_vuln_df <- nspp_fe_df %>%
        left_join(nspp_df, by = 'loiczid') %>%
        group_by(loiczid) %>%
        summarize(n_fe = first(n_fe), n_spp = first(n_spp),
                  f_vuln = sum(nspp_fe == 1) / n_fe, 
                  f_wvuln = mean(1/(nspp_fe^2)),
                  f_red  = mean(nspp_fe),
                  .groups = 'drop')
      f_overred_df <- nspp_fe_df %>%
        left_join(f_vuln_df) %>%
        mutate(f_over = ifelse(nspp_fe > f_red, nspp_fe - f_red, 0)) %>%
        group_by(loiczid) %>% 
        summarize(f_overred = sum(f_over) / sum(nspp_fe),
                  .groups = 'drop')
    # }) ### 7 sec
    
    fe_metrics_df <- f_vuln_df %>%
      full_join(f_overred_df, by = 'loiczid') %>%
      mutate(across(where(is.numeric), .fns = ~round(.x, 6)))
      ### rounding to 6 dec places saves a lot of storage for tifs
  # })
    # ggplot(fe_metrics_df, aes(x = f_vuln, y = f_wvuln)) +
    #   geom_point(alpha = .3) +
    #   geom_abline(slope = 1, intercept = 0)
  
  write_csv(fe_metrics_df, fe_metrics_file)
}

```

## Map functional diversity metrics

Create and save rasters of functional diversity metrics


```{r create rasts}
fe_metrics_df <- read_csv(fe_metrics_file)

n_spp_rast  <- map_to_hcaf(fe_metrics_df, which = 'n_spp')
n_fe_rast   <- map_to_hcaf(fe_metrics_df, which = 'n_fe')
f_vuln_rast <- map_to_hcaf(fe_metrics_df, which = 'f_vuln')
f_wvuln_rast <- map_to_hcaf(fe_metrics_df, which = 'f_wvuln')
f_red_rast  <- map_to_hcaf(fe_metrics_df, which = 'f_red')
f_overred_rast <- map_to_hcaf(fe_metrics_df, which = 'f_overred')

writeRaster(n_spp_rast,   here('_output/functional_entities/n_spp.tif'),   overwrite = TRUE)
writeRaster(n_fe_rast,    here('_output/functional_entities/n_fe.tif'),    overwrite = TRUE)
writeRaster(f_vuln_rast,  here('_output/functional_entities/f_vuln.tif'),  overwrite = TRUE)
writeRaster(f_wvuln_rast, here('_output/functional_entities/f_wvuln.tif'), overwrite = TRUE)
writeRaster(f_red_rast,   here('_output/functional_entities/f_red.tif'),   overwrite = TRUE)
writeRaster(f_overred_rast, here('_output/functional_entities/f_overred.tif'), overwrite = TRUE)
```

Examine distributions of each raster.

```{r histograms}
hist(n_spp_rast)
hist(n_fe_rast)
hist(f_vuln_rast)
hist(f_wvuln_rast)
hist(f_red_rast)
hist(f_overred_rast)
```

Plot each raster.
```{r plot rasters}
plot_map <- function(r, main) {
  var_name <- sym(names(r))
  r_df <- as.data.frame(r, xy = TRUE)
  ### set fill scale limits based on raster values, from 0 to at least 1
  var_lim  <- c(0, max(1, maxValue(r)))
  ggplot(r_df, aes(x, y, fill = !!var_name)) +
    geom_raster() +
    scale_fill_viridis_c(limits = var_lim) +
    coord_sf() +
    theme_void() +
    labs(title = main)
}
plot_map(n_spp_rast,   main = 'Species richness')
plot_map(log10(n_spp_rast) %>% setNames('log10_n_spp'),  main = 'log_10(Species richness)')
plot_map(n_fe_rast,    main = 'Functional richness')
plot_map(log10(n_fe_rast) %>% setNames('log10_n_fe'),  main = 'log_10(Functional richness)')
plot_map(f_vuln_rast,  main = 'Functional vulnerability')
plot_map(f_wvuln_rast, main = 'Weighted functional vulnerability')
plot_map(f_red_rast,   main = 'Functional redundancy')
plot_map(log10(f_red_rast) %>% setNames('log10_f_red'),  main = 'log_10(Functional redundancy)')
plot_map(f_overred_rast, main = 'Functional overredundancy')
```
