---
title: "Map cumulative impacts"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 12)

library(raster)
library(oharac)
library(tidyverse)
library(here)
source(here('common_fxns.R'))
```

# Summary

With impact maps generated for all stressors, means calculated both unweighted and FV-weighted, we can add the impact layers to generate a comprehensive cumulative human impact map.

# Data

See scripts 2, 3a, and 3b for vulnerability mapping scripts, and scripts 4a-4d for impact calculation scripts.

# Methods

## Calculate CHI: unweighted

Sum impacts across all stressor/vulnerability combos.

```{r calc chi by unweighted vuln, fig.height = 3, fig.width = 6}
imp1_fs <- list.files(here('_output/impact_maps/impact_maps_unweighted_all'),
                      pattern = 'mean',
                      full.names = TRUE)

imp1_stack <- stack(imp1_fs)

ocean_a_rast <- raster(here('_spatial/ocean_area_mol.tif'))
### Calculate cumulative impacts
chi_unwt_raw <- calc(imp1_stack, fun = sum, na.rm = TRUE)
chi_unweighted <- chi_unwt_raw %>%
  mask(ocean_a_rast, progress = 'text')

writeRaster(chi_unweighted, here('_output/impact_maps/chi_unweighted.tif'),
            overwrite = TRUE)

### plot with ggplot
chi_unwt_plot <- here('figs/chi_unweighted.png')
if(!file.exists(chi_unwt_plot)) {
  ### unlink(chi_unwt_plot)
  chi_unweighted_df <- chi_unweighted %>%
    as.data.frame(xy = TRUE) %>%
    filter(!is.na(layer))
  
  fill_lbls <- c(0, .1, .25, .5, 1, round(maxValue(chi_unweighted), 2))
  
  p <- ggplot(chi_unweighted_df, aes(x, y, fill = sqrt(layer))) +
    geom_raster() +
    scale_fill_viridis_c(labels = fill_lbls, breaks = sqrt(fill_lbls)) +
    coord_sf() +
    theme_void() +
    labs(title = 'CHI by unweighted mean vulnerability',
         fill  = 'CHI')
  
  ggsave(chi_unwt_plot, height = 4, width = 8, dpi = 300)
}
knitr::include_graphics(chi_unwt_plot)
```

## Calculate CHI: FV-weighted

Sum impacts across all stressor/vulnerability combos.

```{r calc chi by fv-weighted vuln, fig.height = 3, fig.width = 6}
imp2_fs <- list.files(here('_output/impact_maps/impact_maps_fv_weighted'), 
                      pattern = 'mean',
                      full.names = TRUE)

imp2_stack <- stack(imp2_fs)

ocean_a_rast <- raster(here('_spatial/ocean_area_mol.tif'))
### Calculate cumulative impacts
chi_fv_wt_raw <- calc(imp2_stack, fun = sum, na.rm = TRUE)
chi_fv_weighted <- chi_fv_wt_raw %>%
  mask(ocean_a_rast, progress = 'text')

writeRaster(chi_fv_weighted, here('_output/impact_maps/chi_fv_weighted.tif'),
            overwrite = TRUE)

### plot with ggplot
chi_fvwt_plot <- here('figs/chi_fv_weighted.png')
if(!file.exists(chi_fvwt_plot)) {
  ### unlink(chi_fvwt_plot)
  chi_fv_weighted_df <- chi_fv_weighted %>%
    as.data.frame(xy = TRUE) %>%
    filter(!is.na(layer))
  
  fill_lbls <- c(0, .1, .25, .5, 1, round(maxValue(chi_fv_weighted), 2))
  
  p <- ggplot(chi_fv_weighted_df, aes(x, y, fill = sqrt(layer))) +
    geom_raster() +
    scale_fill_viridis_c(labels = fill_lbls, breaks = sqrt(fill_lbls)) +
    coord_sf() +
    theme_void() +
    labs(title = 'CHI by FV-weighted mean vulnerability',
         fill = 'CHI')
  
  ggsave(chi_fvwt_plot, height = 4, width = 8, dpi = 300)
}
knitr::include_graphics(chi_fvwt_plot)
```

## Calculate CHI: unweighted (FE only)

Sum impacts across all stressor/vulnerability combos.

```{r calc chi by unweighted vuln FE spp only, fig.height = 3, fig.width = 6}
imp3_fs <- list.files(here('_output/impact_maps/impact_maps_unweighted_fe_only'), 
                      pattern = '_mean',
                      full.names = TRUE)

imp3_stack <- stack(imp3_fs)

ocean_a_rast <- raster(here('_spatial/ocean_area_mol.tif'))
### Calculate cumulative impacts
chi_unwt_fe_raw <- calc(imp3_stack, fun = sum, na.rm = TRUE)
chi_unweighted_fe <- chi_unwt_fe_raw %>%
  mask(ocean_a_rast, progress = 'text')

writeRaster(chi_unweighted_fe, here('_output/impact_maps/chi_unweighted_fe_only.tif'),
            overwrite = TRUE)

### plot with ggplot
chi_unwt_fe_plot <- here('figs/chi_unweighted_fe_only.png')
if(!file.exists(chi_unwt_fe_plot)) {
  ### unlink(chi_unwt_fe_plot)
  chi_unweighted_fe_df <- chi_unweighted_fe %>%
    as.data.frame(xy = TRUE) %>%
    filter(!is.na(layer))
  
  fill_lbls <- c(0, .1, .25, .5, 1, round(maxValue(chi_unweighted_fe), 2))
  
  p <- ggplot(chi_unweighted_fe_df, aes(x, y, fill = sqrt(layer))) +
    geom_raster() +
    scale_fill_viridis_c(labels = fill_lbls, breaks = sqrt(fill_lbls)) +
    coord_sf() +
    theme_void() +
    labs(title = 'CHI by unweighted mean vulnerability (FE spp only)',
         fill = 'CHI')
  
  ggsave(chi_unwt_fe_plot, height = 4, width = 8, dpi = 300)
}
knitr::include_graphics(chi_unwt_fe_plot)
```

## Plot relative difference in CHI

Difference in CHI calculated from FV-weighted vulnerability relative to unweighted vulnerability (using only spp included in functional entities):
$$\frac{CHI_{FV} - CHI_{unweighted}}{CHI_{unweighted}} \times 100\%$$

```{r fv weighted vs unweighted, fig.height = 4, fig.width = 7}
r <- (chi_fv_weighted - chi_unweighted_fe) / chi_unweighted_fe * 100

r_vals <- values(r); r_vals <- r_vals[!is.na(r_vals)]
r_zlim <- max(abs(quantile(r_vals, .001)), abs(quantile(r_vals, .999)))
values(r)[values(r) > r_zlim] <- r_zlim
values(r)[values(r) < -r_zlim] <- -r_zlim
  
chi_diff_plot <- here('figs/chi_fv_diff_unwt.png')
if(!file.exists(chi_diff_plot)) {
  chi_diff_df <- as.data.frame(r, xy = TRUE) %>%
    filter(!is.na(layer))
  
  map_cols <- hcl.colors(palette = 'Red-Green', n = 15, rev = TRUE)

  p <- ggplot(chi_diff_df, aes(x, y, fill = layer)) +
      geom_raster() +
      scale_fill_gradientn(colors = map_cols, 
                           limits = c(-r_zlim, r_zlim)) +
      coord_sf() +
      theme_void() +
      labs(title = 'CHI by unweighted mean vulnerability (FE spp only)',
           fill = 'CHI')
    
  ggsave(chi_diff_plot, height = 4, width = 8, dpi = 300)
}
knitr::include_graphics(chi_diff_plot)
```

