---
title: "Compare cumulative impacts per species to IUCN status"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(tidyverse)
library(here)
library(RColorBrewer)
library(oharac)
library(caret)
source(here('common_fxns.R'))

```

# Summary

Calculate the cumulative impact across each species range (total and coastal-only).  Relative CHI will be the total CHI (summed across all cells and all stressors) normalized by the total range (total and coastal-only).  The relative CHI is effectively the average CHI per km^2^ across the species' range.

Perform a regression of IUCN extinction risk classification as a function of relative overall CHI and overall range, and then do the same for coastal.

# Data

* Species impact scores from script 5
* IUCN extinction risk scores from IUCN

# Methods

## Read in species impacts for those spp assessed by IUCN; summarize to CHI

Summarize to cumulative impacts by entire range and by coastal range.

Each file has a row for each stressor-vulnerability match; columns are:

* `str_vuln` (chr): name of stressor/vuln combo, as <str>-<vuln>
* `impact_tot` (dbl): sum across all cells of stressor intensity $\times$ vulnerability $\times$ ocean area
* `impact_coastal` (dbl): sum across all coastal cells (≤200m) of stressor intensity $\times$ vulnerability $\times$ ocean area
* `range_tot_km2` (dbl): sum of ocean area of all cells across entire range
* `range_coastal_km2` (dbl): sum of ocean area of all cells across coastal range


```{r}
spp_chi_file <- here_anx('int/spp_chi_summary.csv')

if(!file.exists(spp_chi_file)) {
  spp_imp_fs <- list.files(here_anx('impacts_by_species'), 
                           pattern = '^impacts_.+.csv', full.names = TRUE)
  
  spp_imp_raw_df <- parallel::mclapply(spp_imp_fs, mc.cores = 40,
                                       read_csv, show_col_types = FALSE) %>%
    setNames(basename(spp_imp_fs)) %>%
    bind_rows(.id = 'f')
  
  spp_ranges <- spp_imp_raw_df %>%
    select(f, range_km2, range) %>%
    distinct() %>%
    spread(range, range_km2) %>%
    rename(range_cst_km2 = coastal, range_tot_km2 = full)
  
  spp_intensity_df <- spp_imp_raw_df %>%
    select(f, vuln, stressor, intensity_tot, range) %>%
    spread(range, intensity_tot) %>%
    rename(intensity_tot = full, intensity_cst = coastal)
    
  spp_intens_cv_df <- spp_imp_raw_df %>%
    select(f, vuln, stressor, intensity_cv, range) %>%
    spread(range, intensity_cv) %>%
    rename(intensity_cv_tot = full, intensity_cv_cst = coastal)

  spp_impacts_df <- spp_imp_raw_df %>%
    select(f, vuln, stressor, impact_tot, range) %>%
    spread(range, impact_tot) %>%
    rename(impact_tot = full, impact_cst = coastal)
    
  spp_cv_df <- spp_imp_raw_df %>%
    select(f, vuln, stressor, impact_cv, range) %>%
    spread(range, impact_cv) %>%
    rename(impact_cv_tot = full, impact_cv_cst = coastal)

  spp_impacts_npp_df <- spp_imp_raw_df %>%
    select(f, vuln, stressor, impact_npp_tot, range) %>%
    spread(range, impact_npp_tot) %>%
    rename(impact_npp_tot = full, impact_npp_cst = coastal)
    
  spp_npp_cv_df <- spp_imp_raw_df %>%
    select(f, vuln, stressor, impact_npp_cv, range) %>%
    spread(range, impact_npp_cv) %>%
    rename(impact_npp_cv_tot = full, impact_npp_cv_cst = coastal)

  iucn_worms_lookup <- read_csv(here('_data/iucn_spp', 'iucn_to_worms_match.csv'))

  spp_imp_cv_df <- full_join(spp_impacts_df, spp_cv_df, 
                             by = c('f', 'vuln', 'stressor')) %>%
    full_join(spp_impacts_npp_df, by = c('f', 'vuln', 'stressor')) %>%
    full_join(spp_npp_cv_df, by = c('f', 'vuln', 'stressor')) %>%
    full_join(spp_intensity_df, by = c('f', 'vuln', 'stressor')) %>%
    full_join(spp_intens_cv_df, by = c('f', 'vuln', 'stressor')) %>%
    left_join(spp_ranges, by = 'f') %>%
    ### extract species and map info from filename
    mutate(map_src = str_extract(f, '_iucn_|_am_') %>% str_remove_all('_'),
           iucn_sid = str_extract(f, '[0-9]+') %>% as.integer(),
           spp_from_f = str_remove_all(f, 'impacts_(am|iucn)_|.csv')) %>%
    left_join(iucn_worms_lookup, by = 'iucn_sid') %>%
    mutate(sciname = ifelse(is.na(worms_name), 
                            str_replace_all(spp_from_f, '_', ' '),
                            worms_name)) %>%
    select(sciname, iucn_sid, map_src, vuln, stressor, 
           impact_cst, impact_tot, impact_cv_cst, impact_cv_tot,
           impact_npp_cst, impact_npp_tot, 
           impact_npp_cv_cst, impact_npp_cv_tot,
           intensity_cst, intensity_tot, 
           intensity_cv_cst, intensity_cv_tot,
           range_cst_km2, range_tot_km2) %>%
    distinct() %>%
    filter(range_tot_km2 > 0 & !is.na(impact_tot))
    
  write_csv(spp_imp_cv_df, spp_chi_file)
}
```

### Assemble IUCN Red List categories for species in IUCN maps and AquaMaps maps

``` {r}
iucn_worms_lookup <- read_csv(here('_data/iucn_spp', 'iucn_to_worms_match.csv'),
                              show_col_types = FALSE)

iucn_assessed <- read_csv(here('_data/iucn_spp/iucn_marine_spp_info_2021-3.csv'),
                              show_col_types = FALSE) %>%
  select(iucn_sid, cat, cat_score) %>%
  left_join(iucn_worms_lookup, by = 'iucn_sid') %>%
  mutate(threatened = case_when(cat == 'dd' ~ NA,
                                cat %in% c('lc', 'nt') ~ FALSE,
                                TRUE ~ TRUE),
         atrisk     = case_when(cat == 'dd' ~ NA,
                                cat %in% c('lc') ~ FALSE,
                                TRUE ~ TRUE)) %>%
  select(iucn_sid, iucn_sciname = sciname, sciname = worms_name, cat, cat_score, threatened, atrisk)
iucn_mapped <- read_csv(here('_data/iucn_spp/spp_marine_maps_2021-3.csv'),
                              show_col_types = FALSE)

assessed_in_am <- get_am_spp_info() %>%
  select(am_sid, am_sciname, sciname, iucn_sid) %>%
  inner_join(iucn_assessed %>% 
               select(iucn_sid, cat, threatened, atrisk) %>% distinct(), 
             by = 'iucn_sid') %>%
  select(sciname, cat, threatened, atrisk) %>%
  distinct() %>%
  mutate(map_src = 'am',
         iucn_sid = -1)

assessed_in_iucn_maps <- iucn_assessed %>%
  filter(iucn_sid %in% iucn_mapped$iucn_sid) %>%
  select(sciname, iucn_sid, cat, threatened, atrisk) %>%
  distinct() %>%
  mutate(map_src = 'iucn')

assessed_df <- bind_rows(assessed_in_am, assessed_in_iucn_maps)

```

### Combine IUCN categories with impacts

``` {r}
iucn_impacts_df <- read_csv(spp_chi_file,
                            show_col_types = FALSE) %>%
  mutate(iucn_sid = ifelse(is.na(iucn_sid), -1, iucn_sid)) %>%
  inner_join(assessed_df, by = c('sciname', 'iucn_sid', 'map_src')) %>%
  mutate(iucn_fct = toupper(cat),
         iucn_fct = factor(iucn_fct, levels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD')))

check_df <- iucn_impacts_df %>%
  select(sciname, iucn_sid, map_src) %>%
  distinct() %>%
  group_by(sciname) %>%
  summarize(n = n(),
            map_src = paste(map_src, collapse = ';'))


spp_chi_iucn_clean <- iucn_impacts_df %>%
  filter(stressor == 'cumulative') %>%
  mutate(chi_rel_tot = impact_tot / range_tot_km2,
         chi_rel_cst = impact_cst / range_cst_km2,
         chi_npp_tot = impact_npp_tot / range_tot_km2,
         chi_npp_cst = impact_npp_cst / range_cst_km2) %>%
  filter(!is.na(threatened))
```

## Try some simple tests

### examine CHI by range

```{r}
ggplot(spp_chi_iucn_clean, aes(x = range_tot_km2, y = impact_tot)) +
  geom_point(aes(color = iucn_fct)) +
  geom_abline(slope = 1, intercept = 0, color = 'red', size = .5) +
  # scale_y_log10() +
  theme_minimal()

```

Technically possible for species to have higher total impacts than range, e.g., if high vulnerability (near 1) to multiple very intense stressor (near 1) across most of its range.

Questions:

* Are the spp near the 1:1 line due to cumulative effects, or dominated by a single stressor?  
* there seem to be a few "clusters" of slopes...

## Binomial logistic regression

Perform a binomial logistic regression of threatened status (threatened/not threatened) as a function of mean cumulative impact and range.  Set up data frame to include a `threatened` column, where LC and NT are `FALSE` and VU, EN, and CR are `TRUE`.  Include range as linear rescale (0-1) and square root of range, for both overall range and coastal-only range.

### Prob of threatened = TRUE as function of CHI and range

```{r}
taxa_df <- assemble_worms() %>%
  mutate(taxon = ifelse(phylum == 'chordata' | class == 'cephalopoda', class, phylum)) %>%
  select(species, taxon, phylum, class)

chi_threatened_valid <- spp_chi_iucn_clean %>%
  left_join(taxa_df, by = c('sciname' = 'species')) %>%
  filter(!is.na(threatened)) %>%
  mutate(inv_log_r_tot  = 1/log(range_tot_km2),
         inv_log_r_cst  = 1/log(range_cst_km2),
         inv_sqrt_r_tot = 1/sqrt(range_tot_km2),
         inv_sqrt_r_cst = 1/sqrt(range_cst_km2)) %>%
  mutate(qtile_r = ntile(range_tot_km2, n = 4))

```

``` {r}
taxa <- chi_threatened_valid$taxon %>% unique() %>% sort()

results <- data.frame()

frms <- c(atrisk ~ chi_npp_tot + impact_npp_cv_tot + inv_log_r_tot,
          atrisk ~ chi_npp_cst + impact_npp_cv_cst + inv_log_r_cst,
          atrisk ~ chi_rel_tot + impact_cv_tot + inv_log_r_tot,
          atrisk ~ chi_rel_tot + impact_cv_tot + inv_log_r_tot,
          atrisk ~ chi_rel_tot + inv_log_r_tot,
          atrisk ~ chi_rel_tot + impact_cv_tot,
          atrisk ~ chi_rel_tot,
          atrisk ~ chi_rel_cst + impact_cv_cst + inv_log_r_cst,
          atrisk ~ chi_rel_cst + inv_log_r_cst,
          atrisk ~ chi_rel_cst + impact_cv_cst,
          atrisk ~ chi_rel_cst)

for(t in taxa) { ### t <- taxa[8]
  chi_tx_df <- chi_threatened_valid %>%
    filter(taxon == t) # %>% filter(cat != 'lc')
  
  blr_list <- lapply(frms,
                     FUN = function(f) {
                       mdl <- glm(f, 
                                  data = chi_tx_df,
                                  family = 'binomial') %>%
                         broom::tidy() %>% mutate(formula = deparse(f))
                     })
  
  df <- bind_rows(blr_list) %>%
    mutate(taxon = t)
  
  results <- bind_rows(results, df)
}


x <- results %>%
  filter(term != '(Intercept)')


mdl <- glm(atrisk ~ chi_npp_cst + impact_npp_cv_cst + inv_log_r_cst, 
           data = chi_threatened_valid %>% filter(taxon == 'reptilia'),
           family = 'binomial')
summary(mdl)
```


### Try PCA on indiv stressors then BLR

PCA can eliminate multicollinearity among variables.

```{r spread stressors function}
spread_stressors <- function(df) {
  pca_ready_df <- df %>%
    select(sciname, iucn_sid, stressor, imp_rel_tot, threatened, inv_log_r_tot) %>%
    group_by(stressor) %>%
    ### identify and drop any all-zero-stressor 
    filter(!all(imp_rel_tot == 0)) %>%
    ungroup() %>%
    spread(stressor, imp_rel_tot) %>%
    mutate(across(where(is.numeric), .fns = ~ ifelse(is.na(.x), 0, .x)))
  return(pca_ready_df)
}

prep_for_glm <- function(df, pca_thresh = 0.95) {
  pca_out <- df %>%
    select(-sciname, -iucn_sid, -threatened, -inv_log_r_tot) %>%
    # scale() %>% ### values already between 0-1; scale may lose vulnerability magnitude
    prcomp()
  ### summary(pca_out)

  ### Identify the PC number where the threshold is first met
  cum_var <- ((pca_out$sdev)^2 / sum((pca_out$sdev)^2)) %>% cumsum()
  n_pcs <- which(cum_var > pca_thresh)[1]
  message('Threshold of ', pca_thresh, ' results in ', n_pcs, ' principal components...')
  
  pca_results_df <- data.frame(df %>%
                                 select(threatened, inv_log_r_tot), 
                               pca_out$x[ , 1:n_pcs]) %>%
    mutate(threatened = factor(threatened))
  return(pca_results_df)
}
```

#### model accuracy by taxon

```{r}
spp_impacts_iucn_clean <- iucn_impacts_df %>%
  filter(stressor != 'cumulative') %>%
  mutate(imp_rel_tot = impact_tot / range_tot_km2,
         imp_rel_cst = impact_cst / range_cst_km2) %>%
  filter(!is.na(threatened))

impacts_threatened_valid <- spp_impacts_iucn_clean %>%
  left_join(taxa_df, by = c('sciname' = 'species')) %>%
  filter(!is.na(threatened)) %>%
  mutate(inv_log_r_tot  = 1/log(range_tot_km2),
         inv_log_r_cst  = 1/log(range_cst_km2)) %>%
  select(sciname, iucn_sid, vuln, stressor, threatened, imp_rel_tot, taxon, inv_log_r_tot) %>%
  distinct()

taxa <- impacts_threatened_valid$taxon %>% unique() %>% sort()

results_df <- data.frame()

for(t in taxa) { ### t <- taxa[8]
  message('Processing ', t, '...')
  imp_tx_df <- impacts_threatened_valid %>%
    filter(taxon == t) # %>% filter(cat != 'lc')
  
  pct_thr <- sum(imp_tx_df$threatened) / nrow(imp_tx_df)
  zero_rule <- max(pct_thr, 1 - pct_thr)
  
  ### 
  imp_tx_info <- spread_stressors(imp_tx_df)
  
  imp_tx_pca_df <- prep_for_glm(imp_tx_info)

  tr_ctrl <- trainControl(method = 'repeatedcv', number = 10, repeats = 10)
  ### train the model
  mdl1 <- train(threatened ~ ., 
                data = imp_tx_pca_df,
                method = 'glm', family = 'binomial',
                trControl = tr_ctrl)
  tx_results_df <- data.frame(taxon = t,
                              n_spp = nrow(imp_tx_df),
                              pct_thr = pct_thr,
                              n_pcs = ncol(imp_tx_pca_df) - 2,
                              accuracy = mdl1$results$Accuracy,
                              kappa    = mdl1$results$Kappa,
                              zero_rl  = zero_rule) %>%
    mutate(kappa_0 = (accuracy - zero_rl) / (1 - zero_rl))

  results_df <- bind_rows(results_df, tx_results_df)
}

knitr::kable(results_df %>% arrange(-kappa_0), digits = 4)
```

#### across all spp simultaneously

```{r try blr across all taxa}
pct_thr <- sum(impacts_threatened_valid$threatened) / nrow(impacts_threatened_valid)
zero_rule <- max(pct_thr, 1 - pct_thr)

### 
imp_all_info <- spread_stressors(impacts_threatened_valid)

imp_all_pca_df <- prep_for_glm(imp_all_info)

tr_ctrl <- trainControl(method = 'repeatedcv', number = 10, repeats = 10)
### train the model
mdl2 <- train(threatened ~ ., 
              data = imp_all_pca_df,
              method = 'glm', family = 'binomial',
              trControl = tr_ctrl)
all_results_df <- data.frame(taxon = 'all taxa',
                            pct_thr = pct_thr,
                            n_pcs = ncol(imp_tx_pca_df) - 2,
                            accuracy = mdl2$results$Accuracy,
                            kappa    = mdl2$results$Kappa,
                            zero_rl  = zero_rule) %>%
  mutate(kappa_0 = (accuracy - zero_rl) / (1 - zero_rl))

knitr::kable(all_results_df, digits = 4)

```


