---
title: "Read AquaMaps data"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
am_dir <- '~/git-annex/aquamaps_2021'
```

# Summary

Read in and examine the newest AquaMaps data.

# Methods

## Read in and examine species summary file

```{r}
spp_info <- data.table::fread(file.path(am_dir, 'ver10_2019_speciesoccursum_iucn.csv')) %>%
  janitor::clean_names() %>%
  rename(am_sid = species_id, iucn_sid = iucn_id, comname = f_bname) %>%
  mutate(sciname = tolower(paste(genus, species)))
```
Number of unique species (by AM species ID): `r n_distinct(spp_info$am_sid)`

Number of unique IUCN species (by IUCN species ID): `r spp_info %>% filter(!is.na(iucn_sid)) %>% .$iucn_sid %>% n_distinct()`

## Read in and examine species environmental envelope file

```{r}
env_info <- data.table::fread(file.path(am_dir, 'ver10_2019_hspen.csv')) %>%
  janitor::clean_names() %>%
  rename(am_sid = species_id)

names(env_info)
```

Each parameter (`depth`, `temp`, `salinity`, `prim_prod`, `ice_con`, `oxy`, `land_dist`) has a column for `min`, `pref_min`, `pref_max`, `max` (and `mean` for `depth`).  Additionally, all parameters also have columns for `_yn` (whether that parameter is relevant for that species). 
For each parameter, let's plot distributions of the min, max, and preferred min/max for all spp that depend on that parameter.

```{r examine dists for each parameter}
params <- c('depth', 'temp', 'salinity', 'prim_prod', 'ice_con', 'oxy', 'land_dist')
prm_or <- paste0(params, collapse = '|')
dist_vals <- c('min', 'pref_min', 'mean', 'pref_max', 'max')
env_sum_df <- env_info %>%
  rename(depth_mean = mean_depth) %>%
  select(am_sid, contains(params)) %>%
  gather(param_full, value, -am_sid) %>%
  mutate(param = str_extract(param_full, prm_or),
         dist = str_replace(param_full, paste0('(', prm_or, ')_'), '')) %>%
  select(-param_full) %>%
  spread(dist, value) %>%
  filter(yn == 1) %>%
  select(-yn) %>%
  gather(dist, value, -am_sid, -param) %>%
  mutate(dist = factor(dist, levels = dist_vals),
         value = ifelse(param == 'depth', log10(value + 1), value),
         param = ifelse(param == 'depth', 'log10_depth', param))

write_csv(env_sum_df, file.path(am_dir, 'species_envelope_summary.csv'))

ggplot(env_sum_df) +
  geom_density(aes(x = value, color = dist, ..scaled..), fill = NA) +
  scale_color_viridis_d() +
  facet_wrap(~param, scales = 'free_x') +
  theme_minimal()
```

``` {r examine range of values for each parameter}
env_range_df <- env_sum_df %>%
  mutate(value = ifelse(param == 'log10_depth', 10^value, value)) %>%
  spread(dist, value) %>%
  mutate(mid = ifelse(is.na(mean), (pref_min + pref_max)/2, mean),
         pref_range = pref_max - pref_min,
         max_range  = max - min,
         pref_range = ifelse(param == 'log10_depth', log10(pref_range), pref_range),
         max_range  = ifelse(param == 'log10_depth', log10(max_range), max_range)) %>%
  arrange(pref_range) %>%
  mutate(spp = fct_inorder(am_sid))

ggplot(env_range_df) +
  geom_density(aes(x = pref_range, ..scaled..)) +
  geom_density(aes(x = max_range, ..scaled..), color = 'darkred') +
  facet_wrap(~ param, scales = 'free_x') +
  theme_minimal()
```

## Read in and examine HCAF file

```{r}
hcaf_info <- data.table::fread(file.path(am_dir, 'hcaf_v7.csv')) %>%
  janitor::clean_names() %>%
  mutate(across(where(is.numeric), ~ ifelse(.x == -9999, NA, .x)))

names(hcaf_info)
```

``` {r map vars, cache = TRUE}
rast_base <- raster(ext = extent(c(-180, 180, -90, 90)), resolution = 0.5, crs = '+init=epsg:4326') %>%
  setValues(1:ncell(.))
var_df <- hcaf_info %>%
  select(loiczid, p_water, eez:land_dist, 
         -starts_with(c('elevation', 'area', 'lme')), -ocean_basin)
vars <- names(var_df)[-1]

for(v in vars) { ### v <- vars[1]
  message('plotting map for ', v)
  var_rast <- subs(rast_base, var_df, by = 'loiczid', which = v)
  plot(var_rast, main = v, col = hcl.colors(n = 100))
}
```

## Read and examine species cell occurrences

```{r}
spp_valid <- spp_info %>%
  filter(occur_cells >= 10) %>%
  select(am_sid, sciname)

# csq_loiczid <- hcaf_info %>%
#   select(loiczid, csquare_code) %>%
#   distinct()
# spp_cells <- data.table::fread(file.path(am_dir, 'hcaf_species_native.csv')) %>%
#   janitor::clean_names() %>%
#   left_join(csq_loiczid) %>%
#   select(am_sid = species_id, loiczid, prob = probability)
# 
# write_csv(spp_cells, file.path(am_dir, 'hcaf_species_native_clean.csv'))
spp_cells <- data.table::fread(file.path(am_dir, 'hcaf_species_native_clean.csv')) %>%
  filter(am_sid %in% spp_valid$am_sid)

spp_rich <- spp_cells %>%
  group_by(loiczid) %>%
  summarize(nspp = n_distinct(am_sid), .groups = 'drop')

spp_rich_rast <- subs(rast_base, spp_rich, by = 'loiczid', which = 'nspp')

plot(log10(spp_rich_rast), col = hcl.colors(20))

```

