---
title: "Figure 3: Comparing species and habitat impacts"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 7)

library(terra)
library(sf)
library(oharac)
library(data.table)
library(tidyverse)
library(biscale)  ### for bivariate legend
library(cowplot)
library(here)
source(here('common_fxns.R'))

```

# Summary

Create Fig. 3(?) for manuscript

# Data

Results from scripts in folder `2_process_spp_vuln_and_impacts` and `3_process_hab_vuln_and_impacts`

# Methods

Read in the output rasters for species-level impacts (unweighted); reclassify based on quartile/quintile.  Fig. 1A is bivariate plot of climate vs. non-climate stressors for equal weighted species average, Fig. 1B is bivariate plot of climate vs. non-climate stressors for FV-weighted FE average; Fig. 1C and 1D are %difference plots for each category.

NOTE: dropping cells with fewer than 20 species, mostly in the Arctic, a few in Antarctic.

```{r}
ocean_map <- rast(here('_spatial/ocean_area_mol.tif'))
ocean_df <- as.data.frame(ocean_map, xy = TRUE)

land_sf <- read_sf(here('_spatial/ne_10m_land/ne_10m_land_no_casp.shp')) %>%
  st_transform(st_crs(ocean_map))

nspp_mask <- rast(here('_output/nspp_maps/species_richness.tif'))
nspp_vals <- values(nspp_mask)
# values(nspp_mask)[nspp_vals < 20] <- NA

chi_gps_hab <- list.files(here('_output/cumulative_impact_maps'),
                          pattern = 'habitat_group_chi',
                          full.names = TRUE)
chi_gps_spp <- list.files(here('_output/cumulative_impact_maps'),
                          pattern = 'species_group_chi',
                          full.names = TRUE)
climate_map_hab <- rast(chi_gps_hab[1]) %>%
  mask(nspp_mask)
climate_map_spp <- rast(chi_gps_spp[1]) %>%
  mask(nspp_mask)

nonclimate_map_hab <- rast(chi_gps_hab[2:4]) %>%
  sum(na.rm = TRUE) %>%
  mask(nspp_mask)

nonclimate_map_spp  <- rast(chi_gps_spp[2:4]) %>%
  sum(na.rm = TRUE) %>%
  mask(nspp_mask)
```

### Reclassify

Reclassify values using `ntile` to get quartile values:
```{r reclass eq weight}
cc_qtile_hab <- noncc_qtile_hab <- climate_map_hab ### initialize two new maps
values(cc_qtile_hab) <- ntile(values(climate_map_hab), 4)
values(noncc_qtile_hab) <- ntile(values(nonclimate_map_hab), 4)
# plot(cc_qtile_hab, axes = FALSE, 
#      main = 'Climate impact quintile, habitat avg', col = hcl.colors(4))
# plot(noncc_qtile_hab, axes = FALSE, 
#      main = 'Non-climate impact quintile, habitat avg', col = hcl.colors(4))

cc_hab_df <- as.data.frame(cc_qtile_hab, xy = TRUE) %>%
  setNames(c('x', 'y', 'cc_hab'))
noncc_hab_df <- as.data.frame(noncc_qtile_hab, xy = TRUE) %>%
  setNames(c('x', 'y', 'noncc_hab'))

cc_noncc_hab_df <- dt_join(cc_hab_df, noncc_hab_df, by = c('x', 'y'), type = 'full') %>%
  mutate(bi_class = paste(cc_hab, noncc_hab, sep = '-'))

```

```{r reclass fv weight}
cc_qtile_spp <- noncc_qtile_spp <- climate_map_spp ### initialize two new maps
values(cc_qtile_spp) <- ntile(values(climate_map_spp), 4)
values(noncc_qtile_spp) <- ntile(values(nonclimate_map_spp), 4)
# plot(cc_qtile_spp, axes = FALSE, 
#      main = 'Climate impact quintile, fv wt avg', col = hcl.colors(4))
# plot(noncc_qtile_spp, axes = FALSE, 
#      main = 'Non-climate impact quintile, fv wt avg', col = hcl.colors(4))
```

## Create panels

### FE impact hot/cool spots overlap with hab impact

Compare the FE CHI hot/cool spots with hab - incl Sørensen similarity

```{r panel X bivariate by spp}
cc_spp_df <- as.data.frame(cc_qtile_spp, xy = TRUE) %>%
  setNames(c('x', 'y', 'cc_spp'))
noncc_spp_df <- as.data.frame(noncc_qtile_spp, xy = TRUE) %>%
  setNames(c('x', 'y', 'noncc_spp'))

cc_noncc_spp_df <- dt_join(cc_spp_df, noncc_spp_df, 
                          by = c('x', 'y'), type = 'full') %>%
  mutate(bi_class = paste(cc_spp, noncc_spp, sep = '-'))

spp_hotspots <- cc_noncc_spp_df %>%
  filter(bi_class %in% c('1-1', '4-4')) %>%
  select(x, y, spp_bi = bi_class)
hab_hotspots <- cc_noncc_hab_df %>%
  filter(bi_class %in% c('1-1', '4-4')) %>%
  select(x, y, hab_bi = bi_class)

similarity_df <- spp_hotspots %>%
  full_join(hab_hotspots, by = c('x', 'y')) %>%
  mutate(match = case_when(is.na(hab_bi) ~ 'b',
                           is.na(spp_bi)  ~ 'c',
                           hab_bi == spp_bi ~ 'a',
                           TRUE ~ 'X')) %>%
  group_by(match) %>%
  summarize(n_gp = n()) %>%
  spread(match, n_gp) %>%
  mutate(jaccard = a / (a + b + c),
         sorensen = 2*a / (
           2*a + b + c))
knitr::kable(similarity_df)
```

### Panel A and B: hab-FE climate and non-climate overlap

``` {r make biscale legend}
map_pal_raw <- bi_pal(pal = 'PurpleOr', dim = 4, preview = FALSE)
  
map_pal_mtx      <- matrix(map_pal_raw, nrow = 4, ncol = 4)
map_pal_mtx[3, ] <- colorspace::lighten(map_pal_mtx[3, ], .1)
map_pal_mtx[2, ] <- colorspace::lighten(map_pal_mtx[2, ], .2)
map_pal_mtx[1, ] <- colorspace::lighten(map_pal_mtx[1, ], .3)
map_pal_mtx[ , 3] <- colorspace::lighten(map_pal_mtx[ , 3], .1)
map_pal_mtx[ , 2] <- colorspace::lighten(map_pal_mtx[ , 2], .2)
map_pal_mtx[ , 1] <- colorspace::lighten(map_pal_mtx[ , 1], .3)
map_pal_mtx[1, 1] <- '#ffffee'

map_pal <- as.vector(map_pal_mtx) %>% setNames(names(map_pal_raw))

panel_lgd_raw <- bi_legend(pal = map_pal, dim = 4,
                           x = 'CHI Hab',
                           y = 'CHI FE')
panel_lgd_raw
```

```{r define globe boundary}
globe_bbox <- rbind(c(-180, -90), c(-180, 90), 
                      c(180, 90), c(180, -90), c(-180, -90)) 
globe_border <- st_polygon(list(globe_bbox)) %>%
  st_sfc(crs = 4326) %>%
  st_sf(data.frame(rgn = 'globe', geom = .)) %>%
  smoothr::densify(max_distance = 0.5) %>%
  st_transform(crs = crs(ocean_map))
```

```{r panel A}
cc_spp_hab_df <- dt_join(cc_hab_df, cc_spp_df, by = c('x', 'y'), type = 'full') %>%
  mutate(bi_class = paste(cc_hab, cc_spp, sep = '-')) %>%
  filter(!is.na(cc_spp) & !is.na(cc_hab))

hab_overlap_summary <- cc_spp_hab_df %>%
  group_by(bi_class) %>%
  summarize(ncell = n(),
            pct = n() / nrow(.))
knitr::kable(hab_overlap_summary, digits = 4, 
             col.names = c('hab/FE CHI overlap, climate', '# of cells', '% of area'))

panel_a <- ggplot() +
  ### background for NA cells:
  geom_raster(data = ocean_df, aes(x, y), fill = 'grey85') +
  ### plot data:
  geom_raster(data = cc_spp_hab_df, aes(x, y, fill = bi_class), show.legend = FALSE) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  ### continents:
  geom_sf(data = land_sf,
          fill = 'grey96', color = NA, 
          size = .1) +
  geom_sf(data = globe_border,
          fill = NA, color = 'grey70', 
          size = .1) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_void() +
  theme(axis.text = element_blank(), 
        axis.title = element_blank())
```

```{r panel B}
noncc_spp_hab_df <- dt_join(noncc_hab_df, noncc_spp_df, by = c('x', 'y'), type = 'full') %>%
  mutate(bi_class = paste(noncc_hab, noncc_spp, sep = '-')) %>%
  filter(!is.na(noncc_spp) & !is.na(noncc_hab))

hab_overlap_summary <- noncc_spp_hab_df %>%
  group_by(bi_class) %>%
  summarize(ncell = n(),
            pct = n() / nrow(.))
knitr::kable(hab_overlap_summary, digits = 4, 
             col.names = c('hab/FE CHI overlap, nonclimate', '# of cells', '% of area'))

panel_b <- ggplot() +
  ### background for NA cells:
  geom_raster(data = ocean_df, aes(x, y), fill = 'grey85') +
  ### plot data:
  geom_raster(data = noncc_spp_hab_df, aes(x, y, fill = bi_class), show.legend = FALSE) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  ### continents:
  geom_sf(data = land_sf,
          fill = 'grey96', color = NA, 
          size = .1) +
  geom_sf(data = globe_border,
          fill = NA, color = 'grey70', 
          size = .1) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_void() +
  theme(axis.text = element_blank(), 
        axis.title = element_blank())
```

## Assemble figure

```{r assemble figure}
fig3_file <- here('5_ms_figs/fig3_comparing_hab_vs_spp_chi.png')
panel_a_map <- get_panel(panel_a)
panel_b_map <- get_panel(panel_b)
panel_lgd   <- get_panel(panel_lgd_raw)

```

```{r}
### parameterize legends for easier changing of label positions:
lgd_xy <- c(x = .875, y = .39, h = .144, w = .12)

fig_3 <- ggdraw() +
  ### draw panels:
  draw_plot(panel_a_map,  x = 0.00, y = 0.505, height = 0.49, width = 0.88) +
  draw_plot(panel_b_map,  x = 0.00, y = 0.005, height = 0.49, width = 0.88) +
  ### draw legend:
  draw_plot(panel_lgd, x = lgd_xy['x'], y = lgd_xy['y'], 
            height = lgd_xy['h'], width = lgd_xy['w']) +
  ### Panel labels:
  draw_label('A   Climate', x = 0.002, y = .97, hjust = 0, vjust = 1, 
             size = 9, fontfamily = 'Helvetica', fontface = 'bold') +
  draw_label('B   Non-climate', x = 0.002, y = .47, hjust = 0, vjust = 0, 
             size = 9, fontfamily = 'Helvetica', fontface = 'bold') +
  ### Legend labels:
  draw_label('Species vs. Habitat\nimpacts (quartiles)',
             x = 0.99, y = .54, hjust = 1, vjust = 0, angle = 0,
             size = 7, color = 'black', fontfamily = 'Helvetica', fontface = 'bold') +
  draw_label('Habitat ->', 
             x = lgd_xy['x'], y = lgd_xy['y'] - .005, 
             hjust = 0, vjust = 1, angle = 0,
             size = 6, color = 'black', fontfamily = 'Helvetica', fontface = 'bold') +
  draw_label('Species ->', 
             x = lgd_xy['x'] - .005, y = lgd_xy['y'], 
             hjust = 0, vjust = 0, angle = 90,
             size = 6, color = 'black', fontfamily = 'Helvetica', fontface = 'bold')

ggsave(fig3_file, width = 12, height = 10, units = 'cm', dpi = 300)

knitr::include_graphics(fig3_file)
```

## Examine correlations

```{r set up linear models}
spp_vs_hab_stack <- c(climate_map_hab,
                      nonclimate_map_hab,
                      climate_map_spp,
                      nonclimate_map_spp) %>%
  setNames(c('hab_cc', 'hab_noncc', 'spp_cc', 'spp_noncc'))

### Keep raster values with xy = TRUE, no need for cell IDs
hab_vs_spp_df <- as.data.frame(spp_vs_hab_stack, xy = TRUE)

### generate a LM based on climate stressors
cc_mdl <- lm(spp_cc ~ hab_cc + 0, data = hab_vs_spp_df)
ggplot(hab_vs_spp_df %>% sample_n(10000), aes(x = hab_cc, y = spp_cc)) +
  geom_point() +
  geom_abline(intercept = 0,
              slope = cc_mdl$coefficients[1],
              color = 'red') +
  geom_abline(intercept = 0,
              slope = 1,
              color = 'blue', alpha = .5) +
  coord_fixed() +
  xlim(c(0, 1)) + ylim(c(0, 1))

### generate a LM based on non-climate stressors
noncc_mdl <- lm(spp_noncc ~ hab_noncc + 0, data = hab_vs_spp_df)
ggplot(hab_vs_spp_df %>% sample_n(10000), aes(x = hab_noncc, y = spp_noncc)) +
  geom_point() +
  geom_abline(intercept = 0,
              slope = noncc_mdl$coefficients[1],
              color = 'red') +
  geom_abline(intercept = 0,
              slope = 1,
              color = 'blue', alpha = .5) +
  coord_fixed() +
  xlim(c(0, 1)) + ylim(c(0, 1))
```

$$I_{spp}^{cc} = 0.085 I_{hab}^{cc} + 0 \quad (adj.R^2 = 0.781)$$

$$I_{spp}^{non-cc} = 0.282 I_{hab}^{non-cc} + 0 \quad (adj.R^2 = 0.574)$$


## SI plot - habitat climate vs. non-climate

Use a four-level bivariate color scale to capture spatial overlap of quartiles for climate and non-climate stressors on habitats.

```{r SI fig}

hab_overlap_summary <- cc_noncc_hab_df %>%
  group_by(bi_class) %>%
  summarize(ncell = n(),
            pct = n() / nrow(.))
knitr::kable(hab_overlap_summary, digits = 4, 
             col.names = c('hab CHI overlap', '# of cells', '% of area'))

hab_cc_noncc <- ggplot() +
  ### background for NA cells:
  geom_raster(data = ocean_df, aes(x, y), fill = 'grey85') +
  ### plot data:
  geom_raster(data = cc_noncc_hab_df, aes(x, y, fill = bi_class), 
              show.legend = FALSE) +
  # bi_scale_fill(pal = pal_name, dim = 4) +
  bi_scale_fill(pal = map_pal, dim = 4) +
  ### continents:
  geom_sf(data = land_sf,
          fill = 'grey96', color = NA, 
          size = .1) +
  geom_sf(data = globe_border,
          fill = NA, color = 'grey70', 
          size = .1) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_void() +
  theme(axis.text = element_blank(), 
        axis.title = element_blank())


panel_si_map <- get_panel(hab_cc_noncc)

```



```{r}
### parameterize legends for easier changing of label positions:
lgd_xy <- c(x = .88, y = .40, h = .264, w = .11)

fig_SI <- ggdraw() +
  ### draw panels:
  draw_plot(panel_si_map,  x = 0.00, y = 0, height = 1, width = 0.85) +
  ### draw legend:
  draw_plot(panel_lgd, x = lgd_xy['x'], y = lgd_xy['y'], 
            height = lgd_xy['h'], width = lgd_xy['w']) +
  ### Legend labels:
  ### panel A labels:
  draw_label('Climate v. non-\nclimate impacts\n(quartiles)',
             x = 0.99, y = .70, hjust = 1, vjust = 0, angle = 0,
             size = 7, color = 'black', fontfamily = 'Helvetica', fontface = 'bold') +
  draw_label('Climate ->', 
             x = lgd_xy['x'], y = lgd_xy['y'] - .005, 
             hjust = 0, vjust = 1, angle = 0,
             size = 6, color = 'black', fontfamily = 'Helvetica', fontface = 'bold') +
  draw_label('Non-climate ->', 
             x = lgd_xy['x'] - .005, y = lgd_xy['y'], 
             hjust = 0, vjust = 0, angle = 90,
             size = 6, color = 'black', fontfamily = 'Helvetica', fontface = 'bold')

fig_si_file <- here('5_ms_figs/figSI_map_hab_cc_noncc.png')
ggsave(fig_si_file, width = 12, height = 5, units = 'cm', dpi = 300)

knitr::include_graphics(fig_si_file)
```
