---
title: "prepare Table S2 for manuscript - species inclusion by phylum/class/order"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 7)

library(terra)
library(oharac)
library(data.table)
library(tidyverse)
library(cowplot)
library(here)
source(here('common_fxns.R'))

```

# Summary

Prep table for supplement

# Methods

Load in WoRMS species list.

```{r}
worms_df <- assemble_worms() %>%
  mutate(class = case_when(class == 'actinopterygii' & order == 'acipenseriformes' ~ 'chondrostei',
                           class == 'actinopterygii' ~ 'teleostei',
                           TRUE ~ class))

```

Load in species info; group by phylum, class, order and count species inclusion.  Use prepped data from Fig. 1 plot.

```{r}
all_spp_info <- assemble_spp_info_df()

iucn_sciname_to_id <- all_spp_info %>%
  select(sciname = species, iucn_sid) %>%
  distinct() %>%
  filter(!is.na(iucn_sid))
spp_df <- read_csv(here('5_ms_figs/int/fig1_spp_summary.csv')) %>%
  select(species, iucn_sid) %>%
  distinct() %>%
  left_join(iucn_sciname_to_id, by = 'iucn_sid') %>%
  mutate(species = ifelse(is.na(species), sciname, species)) %>%
  select(-sciname) %>%
  left_join(worms_df, by = 'species') %>%
  unique()

summary_df <- spp_df %>%
  group_by(phylum, class, order, species) %>%
  summarize(subpops = n() - 1) %>%
  mutate(order = ifelse(phylum == 'chordata', order, '-')) %>%
  ### update actinopterygii giga class to classes
  group_by(phylum) %>%
  mutate(n_phl = n_distinct(species)) %>%
  group_by(class) %>%
  mutate(n_cls = n_distinct(species)) %>%
  group_by(order) %>%
  mutate(order = ifelse(n_distinct(species) < 50 & class == 'teleostei', 'other teleostei', order)) %>%
  group_by(order) %>%
  mutate(n_ord = n_distinct(species)) %>%
  group_by(phylum, class, order) %>%
  summarize(n_phl = unique(n_phl),
            n_cls = unique(n_cls),
            n_ord = unique(n_ord),
            subpops = sum(subpops), .groups = 'drop') %>%
  mutate(n_ord = ifelse(order == '-', NA, n_ord),
         subpops = ifelse(subpops == 0, '-', as.character(subpops)))
  
phylum_df <- summary_df %>%
  group_by(phylum, class, n_cls) %>%
  summarize(subpops = sum(as.numeric(subpops), na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(phylum, desc(n_cls)) %>%
  select(phylum, class, n_cls, subpops) %>%
  mutate(subpops = ifelse(subpops == 0, '-', as.character(subpops))) %>%
  distinct()


vert_df <- summary_df %>%
  filter(phylum == 'chordata') %>%
  mutate(class = factor(class, levels = phylum_df$class) %>% fct_drop()) %>%
  arrange(class, desc(n_ord)) %>%
  mutate(order = fct_inorder(order),
         order = fct_relevel(order, 'other teleostei', after = Inf)) %>%
  arrange(class, order) %>%
  select(class, order, n_cls, n_ord, subpops)
```

Create summaries of total count in WoRMS by phylum, class, order.  Note: need to account for "other teleostei" in its own category for the order summary.

```{r}
worms_phyl_sum <- worms_df %>% 
  group_by(phylum) %>% 
  summarize(n_phy_worms = n_distinct(species))
worms_class_sum <- worms_df %>% 
  group_by(class) %>% 
  summarize(n_cls_worms = n_distinct(species))
worms_order_sum <- worms_df %>% 
  mutate(order = ifelse(class == 'teleostei' & !order %in% vert_df$order,
                        'other teleostei', order)) %>%
  group_by(order) %>% 
  summarize(n_ord_worms = n_distinct(species))
```

Finalize tables and write out
```{r}
phylum_out_df <- phylum_df %>%
  left_join(worms_class_sum, by = 'class') %>%
  mutate(repr = paste0(round(n_cls / n_cls_worms * 100, 1), '%'))
vert_out_df <- vert_df %>%
  left_join(worms_order_sum, by = 'order') %>%
  mutate(repr = paste0(round(n_ord / n_ord_worms * 100, 1), '%'))

write_csv(phylum_out_df, here('5_ms_figs/tableS2_spp_by_phylum.csv'))
write_csv(vert_out_df, here('5_ms_figs/tableS2_spp_verts_by_class.csv'))
```

## Representation by phylum and class

```{r}
DT::datatable(phylum_out_df)
```

## Representation by class and order (for vertebrates)

```{r}
DT::datatable(vert_out_df)
```

