---
title: "Set up master rasters"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(sf)
library(fasterize)
library(oharac)
oharac::setup()
source(here('common_fxns.R'))

```

# Summary

Set up rasters for analysis.

* Master rasters at 10 x 10 km and 1 x 1 km resolution Mollweide CRS.
    * 10 x 10 km raster will double as a cell ID raster.
* Ocean area proportion $\in$ (0, 1], 10 km x 10 km Mollweide
    * create global raster from NE ocean at 1 km x 1 km resolution; aggregate up to 10 km x 10 km
* Coastal area proportion $\in$ (0, 1], 10 km x 10 km Mollweide
    * create global raster from buffered land, at 3 nmi distance, at 1 x 1 km resolution; aggregate up to 10 km x 10 km.
    * can also create one at 1 km buffer.
    * do we need inland buffers?
* Using the 10 km x 10 km ocean area raster as a base, create:
    * EEZ raster
    * MEOW raster
    * Bathymetry raster (to clip species to max depth levels)

# Data

Land and ocean shapefiles from Natural Earth.  Free vector and raster map data @ naturalearthdata.com.

# Methods

## Master raster 1 km and 10 km

``` {r}
moll_crs <- '+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m'

rast_base_10km_file <- here('_spatial/rast_base_mol_10km.tif')

if(!file.exists(rast_base_10km_file)) {
  x <- raster() %>%
    setValues(1)

  rast_base_10km <- projectRaster(x, crs = moll_crs, res = 10000, progress="text")

  writeRaster(rast_base_10km, rast_base_10km_file)
}

### disaggregate 10 x 10 km cells to 1 x 1 km cells
rast_base_1km_file <- here_anx('_spatial/rast_base_moll_1km.tif')

if(!file.exists(rast_base_1km_file)) {
  rast_base_10km <- raster(rast_base_10km_file)

  rast_base_1km <- disaggregate(rast_base_10km, fact = 10, progress = 'text')
  
  writeRaster(rast_base_1km, rast_base_1km_file)
}

```

## Ocean area raster

* Reproject NE ocean shapefile to Mollweide.
* Use `fasterize::fasterize` to rasterize the reprojected ocean shapefile to Mollweide CRS with a 1 km resolution.
    * This will be saved on the server, too large for Github.
    * Values will be 1 (ocean) and 0 (not ocean).
* Aggregate the 1 km raster to 10 km cells, taking the sum of ocean cells and divide by 100 to result in values $\in$ (0, 1].  Cells with no ocean presence will result in NA.

```{r}

ocean_1km_file <- here_anx('_spatial/ocean_moll_1km.tif')
# unlink(ocean_1km_file)

if(!file.exists(ocean_1km_file)) {
  message('Processing ocean area at 1 km resolution...')
  ocean_sf <- read_sf(here('_spatial', 'ne_10m_ocean', 'ne_10m_ocean.shp'))
  
  ocean_moll_sf <- ocean_sf %>%
    st_transform(crs = moll_crs)
  
  rast_base_moll_1km <- raster(rast_base_1km_file)
  
  ocean_moll_1km <- fasterize::fasterize(ocean_moll_sf, rast_base_1km, 
                                         field = NULL, ### NULL sets all polygons to 1
                                         background = NA)
    
  writeRaster(ocean_moll_1km, filename = ocean_1km_file)
}
```

```{r}
ocean_10km_file <- here('_spatial', 'ocean_area_mol.tif')
# unlink(ocean_10km_file)

if(!file.exists(ocean_10km_file)) {
  message('Processing ocean area at 10 km resolution...')
  ocean_moll_1km <- raster(ocean_1km_file)
  
  ocean_moll_10km <- aggregate(ocean_moll_1km, fact = 10, fun = sum, progress = 'text')
  
  # rast_base_10km <- raster(rast_base_10km_file)
  # compareRaster(ocean_moll_10km, rast_base_10km)
  # [1] TRUE
  
  ocean_moll_10km <- ocean_moll_10km / 100
  
  writeRaster(ocean_moll_10km, ocean_10km_file)
}

ocean_moll_10km <- raster(ocean_10km_file)
plot(ocean_moll_10km, main = 'ocean area')
```

## Coastal zone raster

* Reproject NE land shapefile to Mollweide.  Buffer to 3 km.
* Use `fasterize::fasterize` to rasterize the reprojected buffered land shapefile to Mollweide CRS with a 1 km resolution.
    * This will be saved on the server, too large for Github.
    * Values will be 1 (land) and 0 (not land).
* Mask with the ocean 1 km raster to include only ocean-based cells that fall within the 3 km buffer.
* Aggregate the 1 km raster to 10 km cells, taking the sum of remaining cells and divide by 100 to result in values $\in$ (0, 1].
* Coastal cells bordering land will contain the proportion of ocean area within 3 km of land.  
    * Verify that this should always be less than or equal to ocean proportion in the cell.

```{r}
coastal_3nmi_buffer_1km_file <- here_anx('_spatial/coastal_3nmi_buffer_moll_1km.tif')
# unlink(coastal_3nmi_buffer_1km_file)

if(!file.exists(coastal_3nmi_buffer_1km_file)) {
  message('Processing 3 nmi coastal buffer at 1 km resolution...')
  land_sf <- read_sf(here('_spatial', 'ne_10m_land', 'ne_10m_land.shp'))
  
  land_moll_sf <- land_sf %>%
    st_transform(crs = moll_crs) 
  
  land_3nmi_moll_sf <- land_moll_sf %>%
    st_buffer(dist = 5556) %>% ### 3 nautical miles = 5.556 km
    st_cast('MULTIPOLYGON')

  rast_base_moll_1km <- raster(rast_base_1km_file)
  
  land_moll_1km <- fasterize::fasterize(land_3nmi_moll_sf, 
                                        rast_base_moll_1km, 
                                        field = NULL, ### NULL sets all polygons to 1
                                        background = NA)
  
  ocean_moll_1km <- raster(here_anx('_spatial/ocean_moll_1km.tif'))

  coastal_3nmi_buffer_1km <- mask(land_moll_1km, ocean_moll_1km, progress = 'text')

  writeRaster(coastal_3nmi_buffer_1km, 
              filename = coastal_3nmi_buffer_1km_file)
}

```

## EEZ raster

``` {r set up EEZ raster}

eez_rast_file <- here('_spatial', 'eez_mol.tif')

if(!file.exists(eez_rast_file)) {
  
  eez_shp_file  <- file.path(here('_spatial/rgns/rgns_mol_1k.gpkg'))
  eez_sf <- read_sf(eez_shp_file)

  eez_df <- eez_sf %>%
    as.data.frame() %>%
    dplyr::select(-geom)
  eez_ids <- eez_df %>%
    select(rgn_id, rgn_name, rgn_ant_id) %>%
    distinct()
  ### rgn_ant_id is same as rgn_id except for Antarctica.  We don't really
  ### need to chunk out Antarctica at this point so just use rgn_id.

  ocean_moll_10km <- raster(ocean_10km_file)
  
  eez_rast <- fasterize::fasterize(eez_sf, ocean_moll_10km,
                                   field = 'rgn_id', 
                                   fun = 'min')
  
  eez_rast_ocean <- eez_rast %>%
    mask(ocean_moll_10km)
  
  writeRaster(eez_rast_ocean, eez_rast_file, overwrite = TRUE)
  
}
eez_rast_ocean <- raster(eez_rast_file)
plot(eez_rast_ocean, main = 'EEZ raster 10 km^2 Mollweide projection')

```

## MEOW raster

Using the MEOW shapefile, we will include all the shapes at the province level and then mask out land using the area raster (for consistency).

``` {r set up MEOW raster}

meow_rast_file <- here('_spatial', 'meow_mol.tif')

if(!file.exists(meow_rast_file)) {
  meow_sf <- read_sf(here('_spatial/rgns/meow_rgns.gpkg'))

  ocean_moll_10km <- raster(ocean_10km_file)

  meow_sf_moll <- meow_sf %>%
    st_crop(st_bbox(c(xmin = -180, ymin = -90, xmax = +180, ymax = +90))) %>%
    st_cast('MULTIPOLYGON') %>%
    smoothr::densify(max_distance = 0.1) %>%
    select(ECO_CODE_X, geom) %>%
    st_transform(crs(ocean_moll_10km))
  
  plot(meow_sf_moll)
  
  rast_base <- area_rast
  
  meow_rast <- fasterize::fasterize(meow_sf_reclass, rast_base, 
                                   field = 'ECO_CODE_X', fun = 'min')
  
  meow_rast <- meow_rast %>%
    mask(area_rast)
    
  ### check whether there are shoreline gaps between MEOW maps and EEZ maps
  # x <- eez_rast
  # x[!is.na(values(meow_rast))] <- NA
  # plot(x)
  ### MEOWs seem to go inland to prevent shoreline gaps! yay
  
  writeRaster(meow_rast, meow_rast_file, overwrite = TRUE)
} else {
  # git_prov(meow_shp_file, 'input')
  # git_prov(meow_rast_file, 'output')
  meow_rast <- raster::raster(meow_rast_file)
}

plot(meow_rast, main = 'MEOW raster 10 km^2 Mollweide')

```


## Bathymetry raster

Using [General Bathymetric Chart of the Oceans (GEBCO) bathymetry raster](https://www.bodc.ac.uk/data/hosted_data_systems/gebco_gridded_bathymetry_data/), identify min depth of ocean cells.  For neritic and coastal cells, leaving as min depth allows for changing depth thresholds if desired.  We will not clip based on anything below 200 m, so can classify these more coarsely.

NOTE: 1 arcmin and 30 arcsec are available; 30 arcsec is more recent but one arcmin should be sufficient for our purposes.  Resample to 6 arc-minute (0.10°) using min depth in each cell.

Additionally, to ensure some minimum coastal cell presence, we will include a buffer based on the 3nm coastal zone from OHI.

Bathymetry is masked to the ocean area raster.  

### Create and view minimum depth raster

First, set up the coastal zone raster.  The original raster is 1/NA for whether a cell falls into the 3 nm coastal buffer, at 934 m resolution Mollweide.  Here, all NA cells are set to zero, non-NA remain 1, so aggregate will result in a proportional coastal zone coverage (so later we can select coastal cells based on some ocean presence threshold if desired).
```{r}
offshore_rast_10km_file <- here('_spatial', 'coastal_mol.tif')

if(!file.exists(offshore_rast_10km_file)) {
  offshore_rast_file <- file.path(dir_bd_anx, 'spatial', 'three_nm_offshore_mol.tif')
  
  x <- raster(offshore_rast_file)
  # values(x)[is.na(values(x))] <- 0
  y <- raster::aggregate(x, fact = 11, fun = sum, na.rm = TRUE, progress = 'text')
  z <- y / 121 ### divide by number of subcells in each supercell
  writeRaster(z, offshore_rast_10km_file, overwrite = TRUE)
}
```

Now we can use the coastal zone buffer to bolster the shallow water species presence to ensure some minimal inclusion even if the bathymetry drops off immediately.  The coastal zone buffer will also be useful when clipping land-based stressor layers.

``` {r set up bathymetry raster}

bathy_rast_file <- here('_spatial', 'bathy_mol.tif')

gebco_rast_file <- file.path(dir_bd_anx, 'spatial/gebco', 'GRIDONE_1D.nc')
gebco_rast_raw  <- raster(gebco_rast_file)

coastal_rast <- raster(offshore_rast_10km_file)

if(!file.exists(bathy_rast_file)) {
  ### The GEBCO raster is elevation, so depth is neg... for min depth, 
  ### aggregate by max (and vice versa)
  bathy_rast_min <- gebco_rast_raw %>%
    raster::aggregate(fact = 6, fun = max, progress = 'text')
  
  bathy_rast_min <- -1 * bathy_rast_min ### depth as positive
    
  rast_base <- coastal_rast
  
  bathy_rast_raw <- bathy_rast_min %>%
    projectRaster(rast_base, 
                  progress = 'text',
                  method = 'ngb')
  
  ### reclassify:
  ### compress cells above zero to sea level;
  ### compress cells below 200 m to rough depth clips
  bathy_rast <- bathy_rast_raw
  values(bathy_rast)[values(bathy_rast_raw) < 0] <- 0
  values(bathy_rast)[values(bathy_rast_raw) > 200] <- 1000
  values(bathy_rast)[values(bathy_rast_raw) > 1000] <- 2000
  values(bathy_rast)[values(bathy_rast_raw) > 2000] <- 3000
  values(bathy_rast)[values(bathy_rast_raw) > 3000] <- 4000
  values(bathy_rast)[values(bathy_rast_raw) > 4000] <- 5000
  values(bathy_rast)[values(bathy_rast_raw) > 5000] <- 6000
  values(bathy_rast)[values(bathy_rast_raw) > 6000] <- 7000
  values(bathy_rast)[values(bathy_rast_raw) > 7000] <- 8000
  
  bathy_rast <- bathy_rast %>%
    mask(area_rast)
    
  writeRaster(bathy_rast, bathy_rast_file, overwrite = TRUE)

  ### set up all masks with a base filled with NAs
  rast_neritic <- rast_shallow <- raster(bathy_rast)
  values(rast_neritic)[values(bathy_rast) <= 200] <- 1
  values(rast_neritic)[!is.na(values(coastal_rast))] <- 1
  values(rast_shallow)[values(bathy_rast) <=  60] <- 1
  values(rast_shallow)[!is.na(values(coastal_rast))] <- 1
  
  writeRaster(rast_neritic, here('_spatial', 'bathy_mol_neritic.tif'),
              overwrite = TRUE)
  writeRaster(rast_shallow, here('_spatial', 'bathy_mol_shallow.tif'),
              overwrite = TRUE)
  
} else {
  # git_prov(bathy_rast_file, 'output')
  bathy_rast <- raster::raster(bathy_rast_file)
}

plot(bathy_rast, main = 'Bathymetry (minimum depth) raster 10 km^2 Mollweide')

```


```{r}
coastal_3nmi_10km_file <- here('_spatial', 'coastal_3nmi_area_mol.tif')
# unlink(coastal_3nmi_10km_file)
if(!file.exists(coastal_3nmi_10km_file)) {
  message('Processing coastal 3 nautical mile buffer area at 10 km resolution...')
  coastal_moll_1km <- raster(coastal_3nmi_buffer_1km_file)
  
  coastal_moll_10km <- aggregate(coastal_moll_1km, fact = 10, fun = sum, progress = 'text')
  
  coastal_moll_10km <- coastal_moll_10km / 100
  
  writeRaster(coastal_moll_10km, coastal_3nmi_10km_file)
}

coastal_moll_10km <- raster(coastal_3nmi_10km_file)
plot(coastal_moll_10km, main = '3nmi coastal area')

x <- ocean_moll_10km - coastal_moll_10km

min(x, na.rm = TRUE)
```

