---
title: "Stressors: targeted fishing by species and cell - finalize stressors per spp"
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_depth: 1
    toc_float: yes
    number_sections: false
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(oharac) ### remotes::install_github('oharac/oharac')
  ### includes some helper functions, e.g., dt_join
library(tidyverse)
library(here)

source(here('common_fxns.R'))
source('stressor_fxns.R') ### in same dir as this Rmd
```

# Summary

Read in rasters (as csvs) of fisheries catch per species per cell.  Normalize by NPP (surface and benthic); identify a reference point; save as .csvs of cell ID and normalized catch. 

# Data

Data from

* Watson, R.A. and Tidd, A.N. (2018) Mapping nearly a century and a half of global marine fishing: 1869 to 2015. Marine Policy 93, 171-177
* Watson, R. (2017) A database of global marine commercial, small-scale, illegal and unreported fisheries catch 1950-2014. Nature Scientific Data 4 (170039).

# Methods

Briefly, here is the process for finalizing catch stressor layers:

* Process catch data for the species:
    * Read in catch per LOICZID cell (for both pelagic and benthic gears)
    * Convert catch per 0.5° cell to catch per km^2^
    * Reproject to Mollweide (either projectRaster or join to a lookup dataframe of LOICZID to Mollweide cell ID)
    * Convert to pelagic and benthic catch per 10 x 10 km cell.
* Divide each processed catch layer (benthic and pelagic) by the appropriate NPP layer (log-transformed)
* Convert to dataframe of cell ID and NPP-normalized catch, and write out.

## Examine NPP layers

Can these be aggregated upward to 0.5 degrees with little loss of information?  Examine mean and variance of aggregated results.  Low variance within an aggregated cell means little information is lost in the aggregation process.

```{r, eval = FALSE}
surface_npp_raw <- raster(here_anx('stressors/fishing/npp',
                                   'Present.Surface.Primary.productivity.Mean.tif'))
# class      : RasterLayer 
# dimensions : 2160, 4320, 9331200  (nrow, ncol, ncell)
# resolution : 0.08333333, 0.08333333  (x, y)
# extent     : -180, 180, -90, 90  (xmin, xmax, ymin, ymax)
# crs        : +proj=longlat +datum=WGS84 +no_defs 
# source     : Present.Surface.Primary.productivity.Mean.tif 
# names      : Present.Surface.Primary.productivity.Mean 
# values     : 6e-05, 0.257881  (min, max)
agg_mean <- aggregate(surface_npp_raw, fact = 6, fun = mean, progress = 'text')
agg_sd   <- aggregate(surface_npp_raw, fact = 6, fun = sd, progress = 'text')
# agg_min  <- aggregate(surface_npp_raw, fact = 6, fun = min, progress = 'text')
# agg_max  <- aggregate(surface_npp_raw, fact = 6, fun = max, progress = 'text')

# agg_range <- agg_max - agg_min
agg_cv <- agg_sd / agg_mean
high_cv <- agg_cv; high_cv[high_cv < 0.1] <- NA

plot(agg_sd, axes = FALSE, col = hcl.colors(20), main = 'std dev')
plot(high_cv, axes = FALSE, col = hcl.colors(20), main = 'coef of var ≥ 0.1')

```
The coastal regions have a relatively high variance, as could be expected.  Because of this, we should probably retain the original resolution.

### Read in processed NPP layers

These layers are created in script `3_process_fishing_bycatch.Rmd`.  The surface NPP raster is the log(x + max(x)/100) transformed surface NPP, normalized by the 99.9th percentile.  The benthic NPP raster is the log-transformed combination of benthic NPP (where non-zero) and export flux from surface NPP, to account for the fact that particularly for deep benthic environments, food is derived from NPP sinking from the surface.  See the indicated script for the derivation of the export flux model.

```{r}
surf_npp  <- raster(here_anx('stressors/fishing/npp', 'log_npp_present_surf_gf_mol.tif'))
benth_npp <- raster(here_anx('stressors/fishing/npp', 'log_npp_present_benth_gf_mol.tif'))

plot(surf_npp, axes = FALSE, main = 'Surface log(NPP)', col = hcl.colors(n = 20))
plot(benth_npp, axes = FALSE, main = 'Benthic log(NPP)', col = hcl.colors(n = 20))
```



### Identify species position in water column

Using the water column position functional trait, separate species into pelagic (surface NPP) and non-pelagic (benthic NPP).

```{r}
watson_am_lookup <- read_csv(here('_setup/stressors/int/watson_am_lookup.csv'))
wcol_pos <- read_csv(here('_data/grouping_traits_post_imputation.csv')) %>%
  dplyr::select(species, wcol)

table(wcol_pos$wcol)
#   ben    bp   pel    rf 
# 28453  3832  4243 14804
### the vast majority of species are not pelagic (92%).  But from script 4b,
### the bulk of catch is not benthic.

# catch_fs <- list.files(here_anx('stressors/fishing/total_catch_by_spp_cell'), full.names = TRUE)
# 
# tot_catch_df <- parallel::mclapply(catch_fs, mc.cores = 20,
#                        FUN = function(f) {
#                          # f <- catch_fs[3]
#                          nm <- str_remove_all(basename(f), 'spp_catch_|.csv') %>%
#                            str_replace_all('[^a-z]+', ' ')
#                          x <- read_csv(f, show_col_types = FALSE)
#                          y <- x %>% summarize(pel_sum = sum(pel_tot_catch),
#                                               ben_sum = sum(ben_tot_catch)) %>%
#                            mutate(spp = nm)
#                        }) %>%
#   bind_rows()
# 
# benth_check <- tot_catch_df %>%
#   left_join(wcol_pos, by = c('spp' = 'species')) %>%
#   group_by(wcol) %>%
#   summarize(ben_tot = sum(ben_sum, na.rm = TRUE),
#             pel_tot = sum(pel_sum, na.rm = TRUE))
# 
# knitr::kable(benth_check)
```

Based on sample of about 16000 spp:

|water col pos | benthic total | pelagic total |
|:-------------|--------------:|--------------:|
|benthic       |   30,861,033  |   19,139,614  |
|benthopelagic |    7,720,392  |   23,181,578  |
|pelagic       |   10,930,387  |   75,876,780  |
|reef          |   13,033,919  |   22,844,775  |
|NA            |   11,217,167  |   19,549,369  |

Catch by benthic and pelagic gear types does not seem well linked to position in water column...

## Process catch layers

* Process catch data for the species:
    * Read in catch per LOICZID cell
    * Convert catch per 0.5° cell to catch per km^2^
    * Reproject to Mollweide (either projectRaster or join to a lookup dataframe of LOICZID to Mollweide cell ID)
    * Convert to catch per 10 x 10 km cell.
* Divide the processed catch layer by the appropriate NPP layer (log-transformed)
* Convert to dataframe of cell ID and NPP-normalized catch, and write out.

```{r set up dataframe of cell_id to loiczid to ocean_a lookup}
loiczid_mol <- raster(here('_spatial/loiczid_mol.tif'))
ocean_a_mol <- raster(here('_spatial/ocean_area_mol.tif'))

cell_id_df <- data.frame(loiczid = values(loiczid_mol),
                         ocean_a_mol = values(ocean_a_mol),
                         cell_id = 1:ncell(ocean_a_mol),
                         pel_npp = values(surf_npp),
                         ben_npp = values(benth_npp))

### using ocean area at half deg generated from Natural Earth
ocean_a_0.5 <- raster(here('_spatial/ocean_area_wgs84_0.5deg.tif'))
ocean_a_hcaf_df <- data.frame(ocean_a_hcaf = values(ocean_a_0.5),
                              loiczid = 1:ncell(ocean_a_0.5))
```

```{r}

spp_hcaf_map_fstem <- here_anx('stressors/fishing/total_catch_by_spp_cell',
                               'am_spp_catch_%s.csv')
spp_mol_map_fstem  <- here_anx('stressors/fishing/npp_norm_catch_by_spp_moll',
                               'am_spp_npp_catch_%s_mol.csv')
spp_vec <- watson_am_lookup$spp %>% unique()

### to cut zero-length files
# x <- list.files(dirname(spp_mol_map_fstem), full.names = TRUE)
# df <- data.frame(f = x, fsize = file.size(x))
# y <- df %>% filter(fsize < 100) %>% .$f
# unlink(y)

### to run only missing files
# x <- list.files(dirname(spp_mol_map_fstem), full.names = TRUE)
# x_spp <- basename(x) %>% 
#   str_remove_all('npp_spp_catch_|_mol.csv') %>%
#   str_replace_all('_', ' ')
# spp_vec <- spp_vec[!spp_vec %in% x_spp]

process_species_to_moll <- function(s) {
  # s <- spp_vec[1]
  spp_mol_map_f <- sprintf(spp_mol_map_fstem, str_replace_all(s, ' +', '_'))
  
  if(!file.exists(spp_mol_map_f)) {
    
    i <- which(spp_vec == s)
    message('Processing NPP-normalized species-total catch map for ', s, 
            ' (', i, ' of ', length(spp_vec), ')')

    ### Read in catch per LOICZID cell; convert catch per 0.5° cell to catch per km^2^
    spp_hcaf_map_f <- sprintf(spp_hcaf_map_fstem, str_replace_all(s, ' +', '_'))
    spp_catch_hcaf <- read_csv(spp_hcaf_map_f, show_col_types = FALSE) %>%
      oharac::dt_join(ocean_a_hcaf_df, by = 'loiczid', type = 'left') %>%
      mutate(pel_per_km2 = pel_tot_catch / ocean_a_hcaf,
             ben_per_km2 = ben_tot_catch / ocean_a_hcaf)
    
    ### "Reproject" to Mollweide by joining to LOICZID-cell_id lookup;
    ### normalize by dividing by log-transformed NPP
    spp_catch_mol <- spp_catch_hcaf %>%
      oharac::dt_join(cell_id_df, by = 'loiczid', type = 'left') %>%
      mutate(pel_sum = pel_per_km2 * ocean_a_mol * 100,
             ben_sum = ben_per_km2 * ocean_a_mol * 100,
             pel_norm = pel_sum / pel_npp,
             ben_norm = ben_sum / ben_npp) %>%
      drop_na()
    
    out_df <- spp_catch_mol %>%
      dplyr::select(cell_id, 
             pel_tot_catch = pel_sum,
             ben_tot_catch = ben_sum,
             pel_norm_catch = pel_norm,
             ben_norm_catch = ben_norm) %>%
      mutate(across(where(is.numeric), ~round(.x, 5)))
    
    write_csv(out_df, spp_mol_map_f)
  } else {
    # message('File ', basename(spp_mol_map_f), ' exists... skipping!')
  }
}

tmp <- parallel::mclapply(spp_vec, mc.cores = 16, FUN = process_species_to_moll)
```

